---
title: "Homework 1 - IND2627"
author:
  - name: Daniel L Santos
    affiliations:
      - name: Department of Industrial Engineering
      - name: PUC-Rio
#format: html
format: pdf
pdf-engine: pdflatex
date: today    
editor: visual
---

## Exercise 1

```{r}
#| echo: true
#| warning: false
#| message: false
#| results: hide
    #Necessary packages for this exercise
    library(tidyverse)
    library(MASS)
```

The <a href="https://www.cs.toronto.edu/~delve/data/boston/bostonDetail.html" target="_blank"> Boston dataset </a> is a dataset derived from information collected by the U.S Census Service concerning housing in the area of Boston Mass. It is composed by

The @tbl-boston shows the first six observations from the dataset.

```{r}
#| echo: false
#| label: tbl-boston
#| tbl-cap: "Boston dataset"
knitr::kable(head(Boston))
```

### Fitting the multiple linear regression

```{r}
    boston_regression_model <- lm(medv ~ lstat + age, Boston)
    boston_regression_model
```

As we can observe, the estimated model function $\hat{Y} = \hat{f}(X)$ is: $$
\hat{\text medv} = -1.03207*{\text lstat} + 0.03454 *{\text age} + 33.22276.
$$ The estimated coefficients are $\hat{\beta}_0 = 33.22276$, $\hat{\beta}_1 = -1.03207$ and $\hat{\beta}_2 = 0.03454$. Notice that the coefficient $\hat{\beta}_1$ is negative, which means that, for a fixed value of *age*, a one-unit increase in *lstat* leads to a decrease in *medv* by $1.03207$ units. On the other hand, for a fixed value of *lstat*, a one-unit increase in *age* leads to an increase in *medv* by $0.03454$ units. Finally, when both *age* and *lstat* are zero, the average value of *medv* will be $\hat{\beta}_0 = 33.22276$.

```{r}
summary(boston_regression_model)
```

```{r}
confint(boston_regression_model)
```


